---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tibble 

[![Build Status](https://travis-ci.org/hadley/tibble.svg?branch=master)](https://travis-ci.org/hadley/tibble) 
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/hadley/tibble?branch=master&svg=true)](https://ci.appveyor.com/project/hadley/tibble) 
[![Coverage Status](https://img.shields.io/codecov/c/github/hadley/tibble/master.svg)](https://codecov.io/github/hadley/tibble?branch=master)

tibble extracts the idea of a __data_frame__ (aka a tibble diff, or tibble for short) from dplyr. As the name suggests a __data_frame__ is a modern reimagining of a data.frame, keeping what time has proven to be effective, and throwing out what is not. In spoken language it's hard to tell the difference between a `data.frame` and `data_frame` so we call the new style tibble dfs (inspired by `dplyr::tbl_df()`), or just tibbles for short.

## Creating tibbles

You can create a tibble from an existing object with `as_data_frame()`:

```{r}
library(tibble)
as_data_frame(iris)
```

You can create a new tibble from vectors that represent the columns with `data_frame()`:

```{r}
data_frame(x = 1:5, y = 1, z = x ^ 2 + y)
```

`data_frame()` is does much less than `data.frame()`: it never changes the type of the inputs (e.g. it never converts strings to factors!), it never changes the names of variables, and it never creates `row.names()`. You can read more about these features in the vignette, `vignette("tibble")`.

You can define a tibble row-by-row with `frame_data()`:

```{r}
frame_data(
  ~x, ~y,  ~z,
  "a", 2,  3.6,
  "b", 1,  8.5
)
```

## Tibbles vs data frames

There are two main differences in the usage of a data frame vs a tibble: printing, and subsetting.

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. Each column gives both the name and its type. This makes it much eaiser to work with large data:

```{r}
library(nycflights13)
flights
```

Tibbles are strict about subsetting. If you try to access a variable that does not exist, you'll get an error:
    
```{r, error = TRUE}
flights$yea
```
    
Tibbles clearly delinerate `[` and `[[`: `[` always returns another tibble, `[[` always returns a vector.
