---
title: "Invariants for subsetting and subassignment"
#output: rmarkdown::word_document
output: rmarkdown::html_vignette
# devtools::load_all(); rmarkdown::render("vignettes/invariants.Rmd", output_format = rmarkdown::md_document(preserve_yaml = TRUE)); system("pandoc vignettes/invariants.md -o vignettes/invariants.html")
vignette: >
  %\VignetteIndexEntry{invariants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.dftbl {
    width: 100%;
    table-layout: fixed;
    display: inline-table;
}

.error pre code {
    color: red;
}

.warning pre code {
    color: violet;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  #eval = FALSE, echo = FALSE,
  comment = "#>"
)

tibble:::set_dftbl_hooks()

options(
  lifecycle_disable_warnings = FALSE,
  lifecycle_verbose_soft_deprecation = TRUE,
  lifecycle_repeat_warnings = TRUE
)
```

This vignette defines invariants for subsetting and subset-assignment for tibbles, and illustrates where their behaviour differs from data frames. 
The goal is to define a small set of invariants that consistently define how behaviors interact. 
Some behaviors are defined using functions of the vctrs package, e.g. `vctrs::vec_slice()`, `vctrs::vec_recycle()` and `vctrs::vec_as_index()`. 
Refer to their documentation for more details about the invariants that they follow.

The subsetting and subassignment operators for data frames and tibbles are particularly tricky, because they support both row and column indexes, both of which are optionally missing.
We resolve this by first defining column access with `[[` and `$`, then column-wise subsetting with `[`, then row-wise subsetting, then the composition of both.

## Conventions

In this article, all behaviors are demonstrated using one example data frame and its tibble equivalent:

```{r setup}
library(vctrs)
library(tibble)

new_df <- function() {
  df <- data.frame(n = 1:4)
  df$c <- letters[5:8]
  df$li <- list(9, 10:11, 12:14, "text")
  df
}

new_tbl <- function() {
  as_tibble(new_df())
}
```

Results of the same code for data frames and tibbles are presented side by side:

```{r show, dftbl = TRUE, dftbl_always = TRUE}
new_df()
```

If the results are identical (after converting to a data frame if necessary), only the tibble result is shown. 

Subsetting operations are read-only.
The same objects are reused in all examples:

```{r ro}
df <- new_df()
tbl <- new_tbl()
```

Where needed, we also show examples with hierarchical columns containing a data frame or a matrix:

```{r setup2}
new_tbl2 <- function() {
  tibble(
    tb = tbl,
    m = diag(4)
  )
}

new_df2 <- function() {
  df2 <- new_tbl2()
  class(df2) <- "data.frame"
  class(df2$tb) <- "data.frame"
  df2
}

df2 <- new_df2()
tbl2 <- new_tbl2()
```

```{r show-compare-2, dftbl = TRUE}
new_df()
```

For subset assignment (subassignment, for short), we need a fresh copy of the data for each test.
The `with_*()` functions (omitted here for brevity) allow for a more concise notation.
These functions take an assignment expression, execute it on a fresh copy of the data, and return the data for printing.
The first example prints what's really executed, further examples omit this output.

```{r with-def, include = FALSE}
with_df <- function(code, verbose = FALSE) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    df <- new_df()
    !!code
    df
  })
  if (verbose) rlang::expr_print(rlang::quo_get_expr(full_code))
  rlang::eval_tidy(full_code)
}

with_tbl <- function(code, verbose = FALSE) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    tbl <- new_tbl()
    !!code
    tbl
  })
  if (verbose) rlang::expr_print(rlang::quo_get_expr(full_code))
  rlang::eval_tidy(full_code)
}

with_df2 <- function(code) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    df2 <- new_df2()
    !!code
    df2
  })
  rlang::eval_tidy(full_code)
}

with_tbl2 <- function(code) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    tbl2 <- new_tbl2()
    !!code
    tbl2
  })
  rlang::eval_tidy(full_code)
}
```

```{r with-demo, dftbl = TRUE}
with_df(df$n <- rev(df$n), verbose = TRUE)
```

## Column extraction

### Definition of `x[[j]]`

`x[[j]]` is equal to `.subset2(x, j)`. 

```{r double-bracket-equivalent-to-subset2, dftbl = TRUE}
df[[1]]
.subset2(df, 1)
identical(df[[2]], .subset2(df, 2))
identical(df[[3]], .subset2(df, 3))
```

NB: `x[[j]]` always returns an object of size `nrow(x)` if the column exists.

`j` must be a single number or a string, as enforced by `vec_as_index(j)`.

```{r double-bracket-requires-scalar-j-index, dftbl = TRUE}
df[[1:2]]
df[[c("n", "c")]]
df[[TRUE]]
df[[mean]]
```

Numeric NA, numeric out-of-bounds (OOB) values, and non-integers throw an error:

```{r double-bracket-j-oob-numeric, dftbl = TRUE}
df[[NA_integer_]]
df[[-1]]
df[[4]]
df[[1.5]]
df[[Inf]]
```

Character OOB access is silent because a common package idiom is to check for the absence of a column with `is.null(df[[var]])`.

```{r double-bracket-j-oob-character, dftbl = TRUE}
df[["x"]]
```

### Definition of `x$name`

`x$name` is equal to `x[["name"]]`.
Unlike data frames, tibbles do not partially match names.

```{r dollar-equivalent-to-subset-pmatch, dftbl = TRUE}
df$l
```

Because `df$x` is rarely used in packages, it can raise a warning:

```{r dollar-j-oob-character, dftbl = TRUE}
df$not_present
```

## Column subsetting

### Definition of `x[j]`.

`j` is converted to an integer vector by `vctrs::vec_as_index(j, nrow(x), names = names(x))`. Then `x[c(j_1, j_2, ..., j_n)]` is equivalent to `tibble(x[[j_1]], x[[j_2]], ..., x[[j_3]])``.
This implies that `j` must be a numeric or character vector, or a logical vector with length 1 or `nrow(x)`.

```{r bracket-j-definition, dftbl = TRUE}
df[1:2]
```

<!-- TODO: what should happen here? -->
If values of `j` are repeated, then the first occurrence of the name is retained.
Subsequent occurrences are renamed with the `"minique"` strategy suggested in [r-lib/vctrs#576](https://github.com/r-lib/vctrs/issues/576).

```{r backet-j-duplication, dftbl = TRUE}
df[c(1, 1)]
```

`nrow(df[j])` equals `nrow(df)`. 

```{r backet-j-empty, dftbl = TRUE}
df[integer()]
```

### Definition of `x[, j]`

`x[, j]` is equal to `x[j]`.
Tibbles do not perform column extraction if `x[j]` would yield a single column.

```{r bracket-drop, dftbl = TRUE}
df[, 1]
df[, 1:2]
```

### Definition of `x[, j, drop = TRUE]`

For backward compatiblity, `x[, j, drop = TRUE]` performs column __extraction__, returning `x[, j][[1]]` when `ncol(x[, j])` is 1.

```{r bracket-always-returns-tibble-drop, dftbl = TRUE}
df[, 1, drop = TRUE]
```

## Row subsetting

### Definition of `x[i, ]`

`i` must be a numeric or logical vector of length `nrow(x)`, or a logical vector of length 1.

```{r bracket-i-wrong-type, dftbl = TRUE}
df[mean, ]
df[list(1), ]
```

Unlike data frames, only logical vectors of length 1 are recycled.
<!-- TODO: should this be an error? -->

```{r bracket-i-recycle, dftbl = TRUE}
df[c(TRUE, FALSE), ]
```

`i` is converted to an integer vector by `vctrs::vec_as_index(i, nrow(x))`.
This implies that OOB values generate warnings:

```{r bracket-i-oob, dftbl = TRUE}
df[10, ]
df["x", ]
```

`x[i, ]` is equivalent to `tibble(vec_slice(x[[1]], i), vec_slice(x[[2]], i), ...)`. 
(It is not defined in terms of `x[[j]][i]` because that definition does not generalise to matrix and data frame columns.)

NB: scalar logicals are recycled, but scalar numerics are not. That makes the `x[NA, ]` and `x[NA_integer_, ]` return different results.

```{r bracket-i, dftbl = TRUE}
df[NA, ]
df[NA_integer_, ]
```

### Definition of `x[i, , drop = TRUE]`

`drop = TRUE` has no effect when not selecting a single row:

```{r backet-i-drop, dftbl = TRUE}
df[1, , drop = TRUE]
```

<!-- TODO: I think this should probably be an error -->

## Row and column subsetting

### Definition of `x[]` and `x[,]`

`x[]` and `x[,]` are equivalent to `x`.

### Definition of `x[i, j]`

`x[i, j]` is equal to `x[i, ][j]` and `x[j][i, ]`.

### Definition of `x[[i, j]]`

`i` must be a numeric vector of length 1.
`x[[i, j]]` is equal to `x[i, ][[j]]` and `x[i, j][[1]]`.

This implies that `j` must be a numeric or character vector of length 1.

NB: `x[[i, j]]` always returns an object of size 1.

## Assignment

### Definition of `x[[j]] <- a`.

`x[[j]] <- a` is equivalent to restoring to the original class after assigning the (potentially recycled) value to `unclass(x)`:

```r
xl <- unclass(x)
xl[[j]] <- vec_recycle(a, nrow(x))
vec_restore(xl, x)
```

```{r double-bracket-assign-definition, dftbl = TRUE}
with_df(df[[1]] <- 0)
with_df(df[[3]] <- 4:1)
with_df2(df2[[1]] <- 0)
with_df2(df2[[2]] <- 4:1)
```

### Corollary

`[[<-` requires scalar `vec_as_index(j)` (with the exception of new columns, see below).

```{r double-bracket-assign-requires-scalar-j-index, dftbl = TRUE}
with_df(df[[1]] <- 0)
with_df(df[[1:2]] <- 0)
with_df(df[[1:3]] <- 0)
with_df(df[["n"]] <- 0)
with_df(df[[c("n", "c")]] <- 0)
with_df(df[[TRUE]] <- 0)
with_df(df[[FALSE]] <- 0)
with_df(df[[NA]] <- 0)
with_df(df[[NA_character_]] <- 0)
with_df(df[[NA_integer_]] <- 0)
```

`x[[j]] <- a` requires `vec_size(a) == nrow(x)` or `vec_size(a) == 1`.

```{r double-bracket-requires-size, dftbl = TRUE}
with_df(df[[1]] <- 1)
with_df(df[[1]] <- 4:1)
with_df(df[[1]] <- 3:1)
with_df(df[[1]] <- 2:1)
```

`[[<-` supports appending a new column at the end.

```{r double-bracket-assign-supports-new, dftbl = TRUE}
with_df(df[["x"]] <- 0)
with_df(df[[4]] <- 0)
with_df(df[[5]] <- 0)
```

`[[<-` supports changing the type of an existing column, also across the principal column types:

- atomic: `df[[1]]` and `df[[2]]`
- list: `df[[3]]`
- data frame: `df2[[1]]`
- matrix: `df2[[2]]`


```{r double-bracket-assign-supports-type-change, dftbl = TRUE}
with_df(df[[1]] <- df[[2]])
with_df(df[[2]] <- df[[3]])
with_df(df[[3]] <- df2[[1]])
with_df2(df2[[1]] <- df2[[2]])
with_df2(df2[[2]] <- df[[1]])
```

`[[<-` supports removing a column by assigning `NULL` to it.

```{r double-bracket-assign-supports-null, dftbl = TRUE}
with_df(df[[1]] <- NULL)
with_df2(df2[[2]] <- NULL)
```

### Columns don't have inner names.

```{r columns-dont-have-names, dftbl = TRUE}
with_df(df[["x"]] <- c(a = 4, b = 3, c = 2, d = 1))[["x"]]
with_df(df[["x"]] <- rlang::set_names(letters[4:1]))[["x"]]
```

### Exception

List columns can have inner names:

```{r list-columns-can-have-names, dftbl = TRUE}
with_df(df[["x"]] <- list(a = 4, b = 3, c = 2, d = 1))[["x"]]
```

### Definition of `x[j, ] <- data.frame(...)`


### Subassignment corollary

Only values of size one can be recycled.

```{r bracket-i-assign-recycle, dftbl = TRUE}
with_df(df[2:3, ] <- df[1, ])
with_df(df[2:3, ] <- df[1:2, ])
with_df(df[2:4, ] <- df[1:2, ])
with_df2(df2[2:4, ] <- df2[1, ])
with_df2(df2[2:4, ] <- df2[2:3, ])
```


### Exception

For compatibility reasons only a warning is issued for indexing beyond the number of rows.
Appending rows right at the end of the existing data is supported, without warning.

```{r bracket-i-oob-num, dftbl = TRUE}
df[5, ]
df[5:7, ]
df[6, ]
with_df(df[5, ] <- df[1, ])
with_df(df[5:7, ] <- df[1, ])
with_df(df[6, ] <- df[1, ])
df[-5, ]
df[-(5:7), ]
df[-6, ]
with_df(df[-5, ] <- df[1, ])
with_df(df[-(5:7), ] <- df[1, ])
with_df(df[-6, ] <- df[1, ])
```

For compatibility reasons, character vectors that represent numbers are supported.

```{r bracket-i-character, dftbl = TRUE}
df[as.character(1:3), ]
df[as.character(-(1:3)), ]
df[as.character(3:5), ]
df[as.character(-(3:5)), ]
df[NA_character_, ]
with_df(df[as.character(1:3), ] <- df[1, ])
with_df(df[as.character(-(1:3)), ] <- df[1, ])
with_df(df[as.character(3:5), ] <- df[1, ])
with_df(df[as.character(-(3:5)), ] <- df[1, ])
with_df(df[NA_character_, ] <- df[1, ])
```


### Definition of `x[j] <- list(...)`.

`x[j] <- a` is equivalent to converting `j` to column indexes or column names and performing `x[[jj]] <- ...` in a loop.

For converting indexes, the following (internal) function is used.

```r
# With logic for defining names of new columns, see below
j <- vec_as_new_col_index(j, x, a)
```

```{r vec-as-new-col-index}
# Existing columns
tibble:::vec_as_new_col_index(2:3, df, list(2, 3))
tibble:::vec_as_new_col_index(-3, df, list(2, 3))
tibble:::vec_as_new_col_index(c(TRUE, FALSE, TRUE), df, list(2, 3))
#
#
# New columns
tibble:::vec_as_new_col_index("x", df, list(y = 0))
tibble:::vec_as_new_col_index(4, df, list(y = 0))
tibble:::vec_as_new_col_index(4, df, list(0))
#
#
# Errors
tibble:::vec_as_new_col_index(NA, df, list(0))
tibble:::vec_as_new_col_index(NA_integer_, df, list(0))
tibble:::vec_as_new_col_index(NA_character_, df, list(0))
tibble:::vec_as_new_col_index(5, df, list(0))
```

For accessing the name or index of a target column, the following helper is used:

```r
index_or_name <- function(j, ja) {
  index <- j[[ja]]
  if (!is.na(index)) index else names(j)[[ja]]
}
```

Below is a simple version of the update loop:

```r
# Create, update or remove, indexed by position
for (ja in seq_along(j)) {
  jx <- index_or_name(j, ja)
  x[[jx]] <- a[[ja]] # with implicit recycling
}
```

Adding, updating and removing columns may occur in arbitrary order.
A robust implementation that avoids index mismatches might look more like this:

```r
a_null <- map_lgl(a, is_null)

# Create or update, indexed by position
for (ja in which(!a_null)) {
  jx <- index_or_name(j, ja)
  x[[jx]] <- a[[ja]] # with implicit recycling
}

# Remove
j_remove <- j[a_null]
if (has_length(j_remove)) x <- x[-j_remove]
```

```{r bracket-assign-def, dftbl = TRUE}
with_df(df[1:2] <- list("x", 4:1))
with_df(df[c("li", "x", "c")] <- list("x", 4:1, NULL))
with_df(df[1:2] <- list(NULL, 4:1))
```

### Corollaries

`NA` indexes are not supported.

```{r bracket-assign-na, dftbl = TRUE}
with_df(df[NA] <- list("x"))
with_df(df[NA_integer_] <- list("x"))
with_df(df[NA_character_] <- list("x"))
```

`[<-` supports changing the type of an existing column, also across the principal column types:

- atomic: `df[[1]]` and `df[[2]]`
- list: `df[[3]]`
- data frame: `df2[[1]]`
- matrix: `df2[[2]]`


```{r bracket-assign-supports-type-change, dftbl = TRUE}
with_df(df[1] <- df[2])
with_df(df[2] <- df[3])
with_df(df[3] <- df2[1])
with_df2(df2[1] <- df2[2])
with_df2(df2[2] <- df[1])
```

Appending columns at the end (without gaps) is supported.
The name of new columns is determined by the LHS, the RHS, or by name repair (in that order of precedence).

```{r bracket-assign-names, dftbl = TRUE}
with_df(df[c("x", "y")] <- tibble("x", x = 4:1))
with_df(df[3:4] <- list("x", x = 4:1))
with_df(df[4] <- list(4:1))
with_df(df[5] <- list(4:1))
```

`x[j]` and `x[[j]]` can be used interchangeably on the LHS or RHS of an assignment:

```{r double-bracket-equivalent-to-unwrap-single-bracket-update, dftbl = TRUE}
with_df(df[[1]] <- df[[3]])
with_df(df[1] <- list(df[[3]]))
with_df(df[[3]] <- df[[1]])
with_df(df[3] <- list(df[[1]]))
with_df2(df2[[1]] <- df[[1]])
with_df2(df2[1] <- list(df[[1]]))
with_df2(df2[[2]] <- df[[3]])
with_df2(df2[2] <- list(df[[3]]))
```

### Definition of `x[j] <- list(a)`.

`x[j] <- a` is equivalent to `x[j] <- rep_along(j, a)` if `is_list(a, 1)`.

```{r bracket-assign-recycles, dftbl = TRUE}
with_df(df[1:2] <- list(0))
with_df(df[1:2] <- list(0, 0))
with_df(df[1:3] <- list(0))
with_df(df[1:3] <- list(0, 0, 0))
```

### Corollary

Only lists of length one on the RHS are recycled.

```{r bracket-assign-recycles-one, dftbl = TRUE}
with_df(df[1:2] <- list(0, 0, 0))
with_df(df[1:3] <- list(0, 0))
```

### Defintiion of `x[i, j] <- a`

`x[i, j] <-` can update only existing columns, it cannot create or remove columns.

```{r bracket-ij-update-existing-only, dftbl = TRUE}
with_df(df[2:3, "n"] <- 1)
with_df(df[2:3, "x"] <- 1)
with_df(df[2:3, "n"] <- NULL)
```

`x[i, j] <-` can't change the data type of existing columns.

```{r bracket-i-data-type, dftbl = TRUE}
with_df(df[2:3, 1] <- df[1:2, 2])
with_df(df[2:3, 2] <- df[1:2, 3])
with_df(df[2:3, 3] <- df2[1:2, 1])
with_df2(df2[2:3, 1] <- df2[1:2, 2])
with_df2(df2[2:3, 2] <- df[1:2, 1])
```

`x[i, j] <-` can't append rows.

```{r append-rows-only-all-columns, dftbl = TRUE}
with_df(df[5, "n"] <- list(0))
```


### Definition of `x[j] <- a`.

`x[j] <- a` is equivalent to `x[j] <- list(a)` if `is_atomic(a)` or `is_null(a)`.

```{r bracket-assign-wraps, dftbl = TRUE}
with_df(df[1] <- 0)
with_df(df[1] <- list(0))
with_df(df[1] <- 4:1)
with_df(df[1] <- list(4:1))
with_df(df[1] <- NULL)
with_df(df[1] <- list(NULL))
```

### Definition of `x[] <- a`.

`x[]` is equivalent to `x[seq_len(ncol(x))] <- a`.

```{r bracket-empty, dftbl = TRUE}
df[]
with_df(df[] <- df)
with_df(df[] <- df[1, ])
with_df(df[] <- df[1:2, ])
with_df(df[] <- df2)
df2[]
with_df2(df2[] <- df2)
with_df2(df2[] <- df2[1, ])
with_df2(df2[] <- df2[1:2, ])
with_df2(df2[] <- df)
```


### Definition of `x[i, j] <- a`

```{r}
with_df(df[2:3, ] <- df[1, ])
with_df(df[0:2, ] <- df[1, ])
with_df(df[0, ] <- df[1, ])
with_df(df[-2, ] <- df[1, ])
with_df(df[-1:2, ] <- df[1, ])
with_df(df[NA_integer_, ] <- df[1, ])
with_df2(df2[NA_integer_, ] <- df2[1, ])
with_df(df[c(FALSE, TRUE, TRUE, FALSE), ] <- df[1, ])
with_df(df[TRUE, ] <- df[1, ])
with_df(df[FALSE, ] <- df[1, ])
with_df(df[NA, ] <- df[1, ])
```

### Definition of `x[[i, j]] <- a`


```{r double-bracket-i-j-assign, dftbl = TRUE}
with_df(df[[1, 1]] <- 0)
with_df(df[1, ][[1]] <- 0)
with_df(df[[1, 3]] <- list(NULL))
with_df(df[1, ][[3]] <- list(NULL))
with_df2(df2[[1, 1]] <- df[1, ])
with_df2(df2[1, ][[1]] <- df[1, ])
with_df2(df2[[1, 2]] <- t(1:4))
with_df2(df2[1, ][[2]] <- t(1:4))
```

`vec_as_index(i)` must be scalar.

```{r double-bracket-requires-scalar-i-index, dftbl = TRUE}
df[[1:2, 1]]
with_df(df[[1:2, 1]] <- 0)
```



`x[[i, j]]` and `x[i, ][[j]]` can be used interchangeably on the LHS or RHS of an assignment for the four principal column types:

- atomic: `df[[1]]`
- list: `df[[3]]`
- data frame: `df2[[1]]`
- matrix: `df2[[2]]`

```{r double-bracket-i-j-assign2, dftbl = TRUE}
with_df(df[[1, 1]] <- df[2, ][[1]])
with_df(df[1, 1] <- list(df[[2, 1]]))
with_df(df[[1, 3]] <- df[2, ][[3]])
with_df(df[1, 3] <- list(df[[2, 3]]))
with_df2(df2[[1, 1]] <- df2[2, ][[1]])
with_df2(df2[1, 1] <- list(df2[[2, 1]]))
with_df2(df2[[1, 2]] <- df2[2, ][[2]])
with_df2(df2[1, 2] <- list(df2[[2, 2]]))
```
