---
title: "Subsetting and subassignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{subassign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.dftbl {
    width: 100%;
    table-layout: fixed;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>"
)

tibble:::set_dftbl_hooks()

options(
  lifecycle_disable_warnings = FALSE,
  lifecycle_verbose_soft_deprecation = TRUE,
  lifecycle_repeat_warnings = TRUE
)
```

```{r setup}
library(tibble)
```

This vignette is an attempt to provide a comprehensive overview over all subsetting and subassignment operations, highlighting where the tibble implementation differs from the data frame implementation.

```{r def}
new_df <- function() {
  df <- data.frame(a = 1:4)
  df$b <- letters[5:8]
  df$cd <- list(9, 10:11, 12:14, "text")
  df
}

new_tbl <- function() {
  as_tibble(new_df())
}
```

Results of the same code for data frames and tibbles are presented side by side:

```{r show, dftbl = TRUE, dftbl_always = TRUE}
new_df()
```

In the following, if the results are identical (after converting to a data frame if necessary), only the tibble result is shown, as in the example below.
This allows to spot differences easier.

```{r show-compare, dftbl = TRUE}
new_df()
```


## Subsetting

Subsetting operations are read-only.
The same objects are reused in all examples:

```{r ro}
df <- new_df()
tbl <- new_tbl()
```

### $

With `$` subsetting, accessing a missing column gives a warning.
Inexact matching is not supported:

```{r dollar, dftbl = TRUE}
df$a
df$"a"
df$a[2:3]
df$cd
df$c
df$d
```

### [[

The `exact` argument is not supported by tibbles.

```{r double-bracket-col, dftbl = TRUE}
df[["a"]]
df[["cd", exact = TRUE]]
df[["cd", exact = FALSE]]
df[["c", exact = TRUE]]
df[["c", exact = FALSE]]
```

With two indexes, a single element is returned.

```{r double-bracket-both, dftbl = TRUE}
df[[2, "a"]]
df[[2, "cd"]]
df[[1:2, "cd"]]
df[[2, "c"]]
df[[1:2, "c"]]
```

Exotic variants like recursive indexing are deprecated for tibbles.

```{r double-bracket-exotic, dftbl = TRUE}
df[[c(1, 2)]]
```

### [

With `[` subsetting, tibbles always return a tibble.
The `drop` argument is supported but has different defaults:

```{r bracket-col-one, dftbl = TRUE}
df["a"]
df["a", drop = FALSE]
df["a", drop = TRUE]
df[1]
df[0]
df[4]
df[NA]
df[NA_character_]
df[NA_integer_]
```

The same examples are repeated for two-dimensional indexing when omitting the row index:


```{r bracket-col-two, dftbl = TRUE}
df[, "a"]
df[, "a", drop = FALSE]
df[, "a", drop = TRUE]
df[, 1]
df[, 0]
df[, 4]
df[, NA]
df[, NA_character_]
df[, NA_integer_]
```

Multiple columns can be queried by passing a vector of column indexes (names, positions, or even a logical vector).
With the latter option, tibbles are a tad stricter:

```{r bracket-col-multi-one, dftbl = TRUE}
df[c("a", "b")]
df[1:2]
df[1:3]
df[1:4]
df[0:2]
df[-1:2]
df[-1]
df[-(1:2)]
df[c(TRUE, TRUE, FALSE)]
df[c(FALSE, TRUE, FALSE)]
df[c(FALSE, TRUE)]
df[c(FALSE, TRUE, FALSE, TRUE)]
```

The same examples are repeated for two-dimensional indexing when omitting the row index:

```{r bracket-col-multi-two, dftbl = TRUE}
df[, c("a", "b")]
df[, 1:2]
df[, 1:3]
df[, 1:4]
df[, 0:2]
df[, -1:2]
df[, -1]
df[, -(1:2)]
df[, c(TRUE, TRUE, FALSE)]
df[, c(FALSE, TRUE, FALSE)]
df[, c(FALSE, TRUE)]
df[, c(FALSE, TRUE, FALSE, TRUE)]
```

Row subsetting works almost identical, but out-of-bounds subsetting is not recommended and may lead to an error in future versions.

```{r bracket-row, dftbl = TRUE}
df[1, ]
df[1:2, ]
df[0, ]
df[5, ]
df[4:5, ]
df[-1, ]
df[-1:2, ]
df[NA, ]
df[NA_integer_, ]
df[c(NA, 1), ]
```

Indexing both row and column works more or less the same, except for `drop`:

```{r bracket-both, dftbl = TRUE}
df[1, "a"]
df[1, "a", drop = FALSE]
df[1, "a", drop = TRUE]
df[1:2, "a"]
df[1:2, "a", drop = FALSE]
df[1:2, "a", drop = TRUE]
df[1, c("a", "b")]
df[1:2, c("a", "b")]
```

Indexes can be omitted altogether, no differences here:

```{r bracket-empty, dftbl = TRUE}
df[]
df[,]
```

## Subassignment

For subassignment, we need a fresh copy of the data for each test.
The `with_*()` functions allow for a more concise notation
(`with_tbl()` omitted here for brevity):

```{r with-df-def}
with_df <- function(code, verbose = FALSE) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    df <- new_df()
    !!code
    df
  })
  if (verbose) rlang::expr_print(rlang::quo_get_expr(full_code))
  rlang::eval_tidy(full_code)
}
```


```{r with-tbl-def, include = FALSE}
with_tbl <- function(code, verbose = FALSE) {
  code <- rlang::enexpr(code)
  
  full_code <- rlang::quo({
    tbl <- new_tbl()
    !!code
    tbl
  })
  if (verbose) rlang::expr_print(rlang::quo_get_expr(full_code))
  rlang::eval_tidy(full_code)
}
```

This function takes an assignment expression and executes it on a fresh copy of the data.
The first example prints what's really executed, further examples omit this output.

```{r with-demo, dftbl = TRUE}
with_df(df$a <- rev(df$a), verbose = TRUE)
```


### $<-

#### Scalars and full length

Assigning a scalar or a full-length vector to a column consistently overwrites existing data or appends a new column at the end. Partial matching doesn't happen:

```{r dollar-assign-scalar, dftbl = TRUE}
with_df(df$a <- 1)
with_df(df$b <- 1)
with_df(df$c <- 1)
with_df(df$cd <- 1)
```

```{r dollar-assign-full, dftbl = TRUE}
with_df(df$a <- 4:1)
with_df(df$b <- 4:1)
with_df(df$c <- 4:1)
with_df(df$cd <- 4:1)
```

#### Recycling

Tibbles allow recycling only for vectors of length 1 or of the same length as the data.

```{r dollar-assign-recycle, dftbl = TRUE}
with_df(df$a <- 1:2)
with_df(df$a <- 1:3)
with_df(df$a <- 1:5)
with_df(df$c <- 1:2)
```



#### Subset assignment

Updating parts of a column extracted by `$` is the responsibility of the column vector.
Tibbles can't control what happens after `$` has returned.

```{r dollar-assign-subset, dftbl = TRUE}
with_df(df$a[1:2] <- 4:3)
with_df(df$b[1:2] <- 4:3)
with_df(df$c[1:2] <- 4:3)
with_df(df$cd[1:2] <- 4:3)
with_df(df$a[1:3] <- 4:3)
with_df(df$a[1:4] <- 4:3)
```

For columns of the stricter `"vctrs_vctr"` class, this class implements the check, which then works identically for data frames and tibbles:

```{r vctrs}
with_df({ df$v = vctrs::new_vctr(1:4); df$v[1:2] <- vctrs::new_vctr(4:3)})
with_df({ df$v = vctrs::new_vctr(1:4); df$v[1:2] <- vctrs::new_vctr(letters[4:3])})
```

### [[<-

```{r double-bracket-assign, dftbl = TRUE}
with_df(df[["a"]] <- 1)
with_df(df[["a"]] <- "x")
with_df(df[["b"]] <- "x")

with_df(df[[2, "a"]] <- 1)
with_df(df[[2, "a"]] <- "x")
with_df(df[[2, "b"]] <- "x")
with_df(df[[2, "cd"]] <- "x")
with_df(df[[1:2, "cd"]] <- "x")
with_df(df[[2, "c"]] <- "x")
with_df(df[[1:2, "c"]] <- "x")

with_df(df[[c(2, 1)]] <- 1)
with_df(df[[c(2, 1)]] <- "x")
with_df(df[[c(1, 2)]] <- "x")
```

### [<-

#### Scalars

```{r bracket-assign-scalar-col, dftbl = TRUE}
with_df(df[1, "a"] <- 1)
with_df(df[1, "b"] <- 1)
with_df(df[1, "c"] <- 1)
with_df(df[1, "cd"] <- 1)
```

#### Full length columns

```{r bracket-assign-full-col, dftbl = TRUE}
with_df(df[, "a"] <- 4:1)
with_df(df[, "b"] <- 4:1)
with_df(df[, "c"] <- 4:1)
with_df(df[, "cd"] <- 4:1)
with_df(df[, "a"] <- 1)
with_df(df[, "b"] <- 1)
with_df(df[, "c"] <- 1)
with_df(df[, "cd"] <- 1)
```

#### Multiple full length columns

```{r bracket-assign-full-multicol, dftbl = TRUE}
with_df(df[, c("a", "b")] <- 4:1)
with_df(df[, c("a", "b")] <- 1)
with_df(df[, c("a", "b")] <- list(4:1, letters[4:1]))
with_df(df[, c("a", "b")] <- list(4:1, 1))
with_df(df[, c("a", "b", "c")] <- list(4:1, letters[4:1]))
with_df(df[, c("a", "b", "cd")] <- list(4:1, letters[4:1]))
```

#### Full length rows

```{r bracket-assign-full-row, dftbl = TRUE}
with_df(df[2, ] <- 1)
with_df(df[2, ] <- "x")
with_df(df[2, ] <- list(1, "x"))
with_df(df[2, ] <- list(1, "x", 2))
with_df(df[2, ] <- list(1, "x", 2, 3))
```

#### Multiple full length rows

```{r bracket-assign-full-multirow, dftbl = TRUE}
with_df(df[2:3, ] <- 1)
with_df(df[2:3, ] <- 1:2)
with_df(df[2:3, ] <- c("x", "y"))
with_df(df[2:3, ] <- list(1:2, c("x", "y")))
with_df(df[2:3, ] <- list(1:2, "x"))
with_df(df[2:3, ] <- list(1, "x", 2))
with_df(df[2:3, ] <- list(1, "x", 2, 3))
```

#### Unspecified

```{r bracket-assign-unspecified, dftbl = TRUE}
with_df(df[] <- 1)
with_df(df[] <- 4:1)
with_df(df[] <- 3:1)
with_df(df[] <- 5:1)
with_df(df[] <- list(1, "x"))
with_df(df[] <- list(4:1, "x", 2))
with_df(df[] <- list(1, "x", 2))
with_df(df[] <- list(1, "x", 2, 3))
with_df(df[] <- df)
with_df(df[,] <- 1)
with_df(df[,] <- 4:1)
with_df(df[,] <- 3:1)
with_df(df[,] <- 5:1)
with_df(df[,] <- list(1, "x"))
with_df(df[,] <- list(4:1, "x", 2))
with_df(df[,] <- list(1, "x", 2))
with_df(df[,] <- list(1, "x", 2, 3))
with_df(df[,] <- df)
```

#### Subset assignment

Due to tibble's default of `drop = FALSE`, updating a portion of a `[` subset is still safe, because tibble is still in control.
Only one example is given here.
(FIXME: It's not safe yet.)

```{r bracket-assign-subset, dftbl = TRUE}
with_df(df["a"][1, ] <- "b")
```
