---
title: "Tibbles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tibbles}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(tibble)
```

Tibbles are a modern take on data frames. They keep the features that have stood the test of time, and drop the features that used to be convenient but are now frustrating (i.e. converting character vectors to factors).

## Creating

`data_frame()` is a nice way to create data frames. It encapsulates best practices for data frames:

  * It never changes an input's type (i.e., no more `stringsAsFactors = FALSE`!).
    
    ```{r}
    data_frame(x = letters)
    ```
    
    This makes it easier to use with list-columns:
    
    ```{r}
    data_frame(x = 1:3, y = list(1:5, 1:10, 1:20))
    ```
    
    List-columns are most commonly created by `do()`, but they can be useful to
    create by hand.
      
  * It never adjusts the names of variables:
  
    ```{r}
    names(data.frame(`crazy name` = 1))
    names(data_frame(`crazy name` = 1))
    ```

  * It evaluates its arguments lazily and sequentially:
  
    ```{r}
    data_frame(x = 1:5, y = x ^ 2)
    ```

  * It never uses `row.names()`. The whole point of tidy data is to 
    store variables in a consistent way. So it never stores a variable as 
    special attribute.
  
  * It only recycles vectors of length 1. This is because recycling vectors of greater lengths 
    is a frequent source of bugs.

## Coercion

To complement `data_frame()`, tibble provides `as_data_frame()` to coerce objects into tibbles. Generally, `as_data_frame()` methods are much simpler than `as.data.frame()` methods, and in fact, it's precisely what `as.data.frame()` does, but it's similar to `do.call(cbind, lapply(x, data.frame))` - i.e. it coerces each component to a data frame and then `cbinds()` them all together. 

`as_data_frame()` has been written with an eye for performance:

```{r}
l <- replicate(26, sample(100), simplify = FALSE)
names(l) <- letters

microbenchmark::microbenchmark(
  as_data_frame(l),
  as.data.frame(l)
)
```

The speed of `as.data.frame()` is not usually a bottleneck when used interactively, but can be a problem when combining thousands of messy inputs into one tidy data frame.

## Tibbles vs data frames

There are two key differences between tibbles and data frames: printing and subsetting.

### Printing

When you print a tibble, it only shows the first ten rows and all the columns that fit on one screen. It also prints an abbreviated description of the column type:
    
```{r}
data_frame(x = 1:1000)
```

You can control the default appearance with options:

* `options(tibble.print_max = n, tibble.print_min = m)`: if there are more 
  than `n` rows, print only the first `m` rows. Use 
  `options(tibble.print_max = Inf)` to always show all rows.

* `options(tibble.width = Inf)` will always print all columns, regardless
   of the width of the screen.

### Subsetting

Tibbles are quite strict about subsetting. `[` always returns another tibble. Contrast this with a data frame: sometimes `[` returns a data frame and sometimes it just returns a single column:
    
```{r}
df1 <- data.frame(x = 1:3, y = 3:1)
class(df1[, 1:2])
class(df1[, 1])

df2 <- data_frame(x = 1:3, y = 3:1)
class(df2[, 1:2])
class(df2[, 1])
```

To extract a single column use `[[` or `$`:

```{r}
class(df2[[1]])
class(df2$x)
```

Tibbles are also stricter with `$`. Tibbles never do partial matching, and will throw an error if the column does not exist:

```{r, error = TRUE}
df <- data.frame(abc = 1)
df$a

df2 <- data_frame(abc = 1)
df2$a
```
